Composing data visualizations

### 2020-04-25 wip-compost/paper.tex
Finish intro, start DSL section

\NeedsTeXFormat{LaTeX2e}
\documentclass{jfp}

@@

\begin{document}
\maketitle[f]

@@

\section{Introduction}
Let's say we want to create the two charts in Figure~\ref{fig:charts}. The chart on the left is
a bar chart, except that it shows two different values for each bar. The chart on the
right is a line chart, except that it also highlights two parts of the timeline with two different
colors.

There is a plenty of libraries that can draw bar charts and line charts, but adding those extra
features will only be possible if the author already thought about your exact scenario. For
example, Google Charts supports the left chart (it is called Dual-X Bar Chart) but there is no
way for adding a background, let alone based on the X value. The alternative is to use a more
low-level library such as D3. In D3 you construct the chart piece by piece, but then you have to
tediously transform your values to coordinates in pixels. For scientific data visualizations,
you could use an implementation of the Grammar of Graphics such as ggplot2. This lets you
map data to geometric objects (such as points, bars, and lines) and their visual properties
(X and Y coordinate, shape and color). However, the number of options is still somewhat limited.

To a functional programmer, this situation is disappointing. We should be able to
compose a variety of charts from a small number of primitives. The shapes should be defined
not in terms of pixels, but numbers of seats or exchange rates. We should also be able to easily
derive high-level abstractions, such as a bar chart, from those low-level primitives.

\begin{figure}
  \includegraphics[scale=0.57]{figures/charts}
  \vspace{0.25em}
  \caption{Two charts about the UK politics: Comparison of election results from 2017 and 2019 (left)
    and GBP-USD exchange rate with highlighted areas before and after the 23 June 2016 Brexit vote.}
  \label{fig:charts}
\end{figure}

In this paper, we...

@@

\section{Domain specific language}
\label{sec:intro}
\subsection*{Domain-specific values}

All values that are provided when constructing charts are from the chart domain. For example,
when visualizing the population of different countries, X values will be categorical values
such as \strf{United Kingdom} or \strf{Czech Republic} and Y values will be continuous values
such as $\num{66.04}$ million or $\num{10.58}$ million. Compost will automatically map the value
from a domain value to a pixel value. The user does not need to do any rescaling to obtain
value in pixels. This is akin to most high-level charting libraries, but in contrast with some
more flexible libraries like, say, D3.
Compost supports only two-dimensional charts and values can be either categorical (such as
different countries) or continuous (such as population size). A value needs to determine an
exact position in the space available for the chart. For continuous values, this simply means
applying a linear transformation. For categorical values, the situation is more difficult. For % example,
say we are drawing a bar chart with population of three different countries with countries
on the X axis and population on the Y axis. The scale of the X axis is categorical and
Compost divides the available space into three ranges of equal width. However, then \strf{United % Kingdom}
or \strf{Czech Republic} does not determine an exact position but instead a range. To determine
a unique position, we need to attach a value between $\num{0}$ and $\num{1}$ to the
categorical value which identifies a relative position in the available range. This is
illustrated in Figure~\ref{fig:scales}.
\includegraphics[scale=1,trim={0cm 6.5cm 6cm 0cm},clip]{values.pdf}
\caption{Values on a continous scale and categorical scale.}
\label{fig:scales}
More formally, we write $c$ for categorical values such as countries (represented as country
names using strings), $r$ for ratio between $0$ and $1$ and we write $n$ for numerical continuous
values with any range. A value $v$ used in a Compost chart is defined as:
v & = & \kvd{cat}~c, r \lsep \kvd{cont}~n
A value $v$ that is used to create Compost charts can be categorical, written as \kvd{cat} and
consisting of the category and a ratio that determines point in the space allocated for the
cateogry. A continous value, written as \kvd{cont}, consists of a number and will be mapped to
a location based on the implicitly obtained re-scaling.

@@

 ~(\kvd{cat}~\strf{UK}, \num{0}), (\kvd{cont}~\num{00.00}), (\kvd{cat}~\strf{UK}, \num{0}), % (\kvd{cont}~\num{66.04}),\\
\hspace{5.9em}(\kvd{cat}~\strf{UK}, \num{1}), (\kvd{cont}~\num{66.04}), (\kvd{cat}~\strf{UK}, % \num{1}), (\kvd{cont}~\num{00.00}),\\

@@

 ~(\kvd{cat}~\strf{CZ}, \num{0}), (\kvd{cont}~\num{00.00}), (\kvd{cat}~\strf{CZ}, \num{0}), % (\kvd{cont}~\num{10.58}),\\
\hspace{5.9em}(\kvd{cat}~\strf{CZ}, \num{1}), (\kvd{cont}~\num{10.58}), (\kvd{cat}~\strf{CZ}, % \num{1}), (\kvd{cont}~\num{00.00})\\

@@

used for mapping domain-specific values to pixels. While doing this, it also checks for consistency % --

@@

the next section. The \kvd{axis} and \kvd{roundScale} constructs could be defined as derived % constructs,
but it is easier to consider them as primitives for now. The \kvd{axis} primitive simply takes the % shape
$s$ and draws an axis around the contents of $s$, using the inferred scale of $s$ to determine % values

@@

\quad(\kvd{explicitScale}_x~(\kvd{categorical % \strf{CZ},\strf{MN},\strf{UK}})~\ident{populationBar})))

@@

bar of a bar chart and then the two values could be $\kvd{cat}~\strf{One}, 0$ and % $\kvd{cat}~\strf{One}, 1$

@@

\quad (\kvd{nest}_y~(\kvd{cat}~\strf{bottom-chart}, \num{0}), (\kvd{cat}~\strf{bottom-chart}, % \num{1}),

@@

We then nest the (continuous) Y axes using the \kvd{nest} primitive and, at the same time, % implicitly

@@

\item We could define a type system to catch categorical/continuous value mismatch in a single % scale.

@@

\item There might be other kinds of scales - for example, color scale (can have meaning in scatter % plot)
 or a scale for secondary markers (like sizes of bubbles in a bubble chart). We could really say % that

>>>>

\NeedsTeXFormat{LaTeX2e}
\documentclass{jfp}
\usepackage{xcolor}

@@

\definecolor{kvdclr}{rgb}{0.5,0.0,0.5}
\definecolor{fkvdclr}{rgb}{0.0,0.0,0.6}
\definecolor{urlclr}{rgb}{0.0,0.0,0.8}
\definecolor{idclr}{rgb}{0.0,0.0,0.0}
\definecolor{numclr}{rgb}{0.0,0.4,0.0}
\definecolor{strclr}{rgb}{0.6,0.2,0.0}

\newcommand{\lsep}{\;\;|\;\;}
\newcommand{\num}[1]{\textcolor{numclr}{#1}}
\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
\newcommand{\strf}[1]{\textnormal{\textcolor{strclr}{\sffamily #1}}}
\newcommand{\ident}[1]{\textnormal{\textcolor{idclr}{\sffamily #1}}}
\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}
\newcommand{\fkvd}[1]{\textnormal{\textcolor{fkvdclr}{\sffamily #1}}}
\newcommand{\narrow}[1]{\hspace{-0.6em}#1\hspace{-0.6em}}
\newcommand{\urrl}[1]{\textnormal{\textcolor{urlclr}{\url{#1}}}}
\begin{document}
\maketitle[f]

@@

\section{Introduction}
Let's say we want to create the two charts in Figure~\ref{fig:charts}. The chart on the left is
a bar chart that shows two different values for each bar. The chart on the right consists of two
line charts that share the X axis and highlight two parts of the timeline with two different colors.

There is a plenty of libraries that can draw bar charts and line charts, but adding those extra
features will only be possible if the author already thought about your exact scenario. For
example, Google Charts supports the left chart (it is called Dual-X Bar Chart) but there is no
way for adding a background, or sharing an axis between charts. The alternative is to use a more
low-level library such as D3. In D3 you construct the chart piece by piece, but then you have to
tediously transform your values to coordinates in pixels yourself. For scientific plots,
you could use an implementation of Grammar of Graphics such as ggplot2, where a chart is
a mapping from data to geometric objects (such as points, bars, and lines) and their visual
properties (X and Y coordinate, shape and color). However, the range of charts that can be
created using this systematic approach is still somewhat limited.

What would an elegant functional approach to data visualization look like? A functional programmer
would want a domain-specific language that has a small number of primitives; allow us to define
high level abstractions such as a bar chart and its basic building blocks are expressed in terms
of domain values such as the exchange rate, rather than pixels.

\begin{figure}[h]
  \includegraphics[scale=0.57]{figures/charts}
  \vspace{0.25em}
  \caption{Two charts about the UK politics: Comparison of election results from 2017 and 2019 (left)
    and GBP-USD exchange rate with highlighted areas before and after the 23 June 2016 Brexit vote.}
  \label{fig:charts}
\end{figure}

As is often the case with domain-specific languages, finding the right primitives is more of an art
than science. For this reason, I present my answer -- a library named Compost -- as a functional pearl.
I hope to convince the reader that Compost is elegant and I illustrate this with a wide range
of examples. Compost has a number of specific desirable properties:

\begin{itemize}
\item Charts are composed from a small number of primitive building blocks using a small number of
  combinators. In particular, concepts such as bar charts, line charts or charts with aligned
  axes are expressed in terms of more basic concepts.
\item The primitives are specified in domain terms. When drawing a line, the value of an Y coordinate
  is an exchange rate of 1.36 USD/GBP, not 137 pixels from the bottom.
\item Most common chart types can be easily captured as high level abstractions, but there is an
  elengant way of creating a majority of more interesting custom charts.
\item The approach can easily be extended to creating web-based charts that involve animations
  or interaction with the user.
\end{itemize}
%
The presentation in this paper focuses on explaining the primitives and combinators of the
domain-specific language. I outline the structure of an implementation, but omit the details. Filling
those in requires careful thinking about geometry and projections, but there are no unexpected
surprises. A complete F\# implementation, including the examples used in this paper, is available
at: \urrl{http://github.com/compostjs}.

\section{Composing visualization primitives}
I will introduce individual features of the Compost library gradually. The first important aspect of
Compost is that properties of shapes are defined in terms of domain-specific values. I first
explain what this means and then use domain-specific values to specify the core part of the
UK election results bar chart.

\subsection{Domain-specific values}

In the election results chart in Figure~\ref{fig:charts} (left), the X values are categorical
values representing the political parties such as \strf{Conservative} or \strf{Labour}. The
Y values are numerical values representing the number of seats won such as $\num{365}$ MPs.
When creating data visualizations, those are the values that the user needs to specify. This is
akin to most high-level charting libraries such as Google Charts, but in contrast with more
flexible libraries like D3.

Our design focuses on two-dimensional charts with X and Y axes. Values mapped to those axes
can be either categorical (such as different political parties or countries) or continuous
(such as number of votes or exchange rates). The mapping from categorical and continuous values
to exact positions on the chart is done automatically. For continuous values, this simply means
applying a linear transformation. For categorical values, the mapping is more difficult.

For example, in the UK election results chart, the X axis is categorical. The library automatically
divides the available space between the six categorical values (political parties). The value
\strf{Green} does not determine an exact position on the axis, but rather a range. To determine
an exact position, we also need to attach a value between $\num{0}$ and $\num{1}$ to the
categorical value. This identifies a relative position in the available range.

\begin{figure}
\includegraphics[scale=1,trim={0cm 6.5cm 6cm 0cm},clip]{figures/values.pdf}
\caption{On a continuous scale (above), an exact position is determined by a number.
  On a categorical scale (below), an exact position is determined by the category and a
  numerical ratio from 0 to 1.}
\label{fig:scales}
\end{figure}

Figure~\ref{fig:scales} illustrates the two kinds of values using the axes from the UK
election results chart. More formally, we define a value $v$ as either a continuous value
$\kvd{cont}~n$ containing any number $n$ or a categorical value $\kvd{cat}~c, r$, consisting
of a categorical value $c$ (implemented as a string) and a ratio $r$ between $0$ and $1$:
%
\begin{equation*}
\begin{array}{rcl}
v & = & \kvd{cat}~c, r \lsep \kvd{cont}~n
\end{array}
\end{equation*}

\vspace{-2.5em}
\subsection{Visualization primitives}

\begin{equation*}
\begin{array}{rcl}
s & = & \kvd{overlay}~[\,s_1, \ldots, s_n\,]\\
 & | & \kvd{text}~\gamma, v_x, v_y, t\\
 & | & \kvd{line}~\gamma, v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}\\
 & | & \kvd{fill}~\gamma, v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}\\
 & | & \kvd{axis}_{x/y} s
\end{array}
\end{equation*}

xx

\begin{equation*}
\begin{array}{l}
  \kvd{axis}_x~(\kvd{axis}_y~(\kvd{overlay}~[\\
  \quad \fkvd{for}~\ident{party}, \ident{clr17}, \ident{clr19}, \ident{mp17}, \ident{mp19}~\fkvd{in}~\ident{elections}~\rightarrow\\
  \quad\quad \kvd{padding}~0,10,0,10,\,\kvd{overlay}~[\\
  \quad\quad\quad \kvd{fill}~\ident{clr},~(\kvd{cat}~\ident{party}, \num{0}), (\kvd{cont}~\num{0}), (\kvd{cat}~\ident{party}, \num{0}), (\kvd{cont}~\ident{mp17}),\\
  \quad\quad\quad \hspace{2.65em}           ~(\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\ident{mp17}), (\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\num{0})\\
  \quad\quad\quad \kvd{fill}~\ident{clr},~(\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\num{0}), (\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\ident{mp19}),\\
  \quad\quad\quad \hspace{2.65em}           ~(\kvd{cat}~\ident{party}, \num{1}), (\kvd{cont}~\ident{mp19}), (\kvd{cat}~\ident{party}, \num{1}), (\kvd{cont}~\num{0})~~]\\
  ]\,)\,)
\end{array}
\end{equation*}

\subsection{Inferring scales and projections}

\section{Fancy scale features}

\section{Definining abstractions}

\section{Interactive charts}

\section{Implementation structure}

@@

\section{Domain-specific language}
\label{sec:intro}
\subsection*{Domain-specific values}

@@

 ~(\kvd{cat}~\strf{UK}, \num{0}), (\kvd{cont}~\num{00.00}), (\kvd{cat}~\strf{UK}, \num{0}), (\kvd{cont}~\num{66.04}),\\
\hspace{5.9em}(\kvd{cat}~\strf{UK}, \num{1}), (\kvd{cont}~\num{66.04}), (\kvd{cat}~\strf{UK}, \num{1}), (\kvd{cont}~\num{00.00}),\\

@@

 ~(\kvd{cat}~\strf{CZ}, \num{0}), (\kvd{cont}~\num{00.00}), (\kvd{cat}~\strf{CZ}, \num{0}), (\kvd{cont}~\num{10.58}),\\
\hspace{5.9em}(\kvd{cat}~\strf{CZ}, \num{1}), (\kvd{cont}~\num{10.58}), (\kvd{cat}~\strf{CZ}, \num{1}), (\kvd{cont}~\num{00.00})\\

@@

used for mapping domain-specific values to pixels. While doing this, it also checks for consistency --

@@

the next section. The \kvd{axis} and \kvd{roundScale} constructs could be defined as derived constructs,
but it is easier to consider them as primitives for now. The \kvd{axis} primitive simply takes the shape
$s$ and draws an axis around the contents of $s$, using the inferred scale of $s$ to determine values

@@

\quad(\kvd{explicitScale}_x~(\kvd{categorical \strf{CZ},\strf{MN},\strf{UK}})~\ident{populationBar})))

@@

bar of a bar chart and then the two values could be $\kvd{cat}~\strf{One}, 0$ and $\kvd{cat}~\strf{One}, 1$

@@

\quad (\kvd{nest}_y~(\kvd{cat}~\strf{bottom-chart}, \num{0}), (\kvd{cat}~\strf{bottom-chart}, \num{1}),

@@

We then nest the (continuous) Y axes using the \kvd{nest} primitive and, at the same time, implicitly

@@

\item We could define a type system to catch categorical/continuous value mismatch in a single %scale.

@@

\item There might be other kinds of scales - for example, color scale (can have meaning in scatter plot)
 or a scale for secondary markers (like sizes of bubbles in a bubble chart). We could really say that
